---
title: "Figure 1 - *huva*"
author: "Bonaguro et al."
output: 
  html_document:
    code_folding: show
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      smooth_scroll: no
    number_sections: true
editor_options: 
  chunk_output_type: inline
---

This Rmarkdonwn document can be used to reproduce the panels in figure 1 of the manuscript: *huva: human variation in gene expression as surrogate for gene function*.
To run this script without any problem of dependencies or conflicts in the installation we suggest to use the docker image we provide (see README file). 

# Figure 1 (Supp. Figures S1,S2)
The first figure provide the reader with a benchmark of the *huva* package, we measured here the time required for the computation of the 4 *huva experiments*. 
We performed the analysis on standard workstation hardware (Intel Core I7-8700K @3.70GHz, 32 GB RAM memory, Windows 10 Pro version 2004).
Considering the nature of the analysis we expect this not to produce the same result on every system, nevertheless we expect that similar hardware will give comparable results.

To run the huva package we do not provide any minimum requirements, we expect any system able to run the requirements of the *huva* package should provide sufficient resources for huva analysis.

## Loading required packages
```{r}
library(huva)
library(huva.db)
library(tictoc)
library(gridExtra)
library(ggplot2)
library(grid)
```

## Costum Functions
```{r}
plot <- function(vector, title="", fill, bins=30) {
  gene_df <- as.data.frame(vector)
colnames(gene_df) <- "time"

p1 = ggplot(aes(x = time), data = gene_df) +
  geom_histogram(fill = fill, color = "black", aes(y=..density..), alpha=0.5, bins = bins) +
  labs(x = "", y = "density", title = (title)) + 
  theme_bw() + theme(aspect.ratio = 1)

p2 = ggplot(aes(x = "", y = time), data = gene_df) +
  stat_boxplot(geom ='errorbar', width = 0.4) +
  geom_boxplot(outlier.colour = "black") +
  coord_flip() +
  labs(x = "", y = "time (s)") +
  theme_void() + theme(aspect.ratio = 1/6)


grid.draw(rbind(ggplotGrob(p1), ggplotGrob(p2), size = "first"))
}
```

## Calculate the benchmark
Here we are going to calculate the average time required for each of the *huva* experiments, the package tictoc will be use to keep track of the time required for each experiment

### Prepare a container
```{r}
bench <- list()
```

### Setting seed
Setting the seed will allow to always capture the same genes in the simulation
```{r}
set.seed(91)
```

### *huva GOI experiment*
Here we will calculate the time required for the run of an *huva GOI experiment*
```{r, echo = T, results = 'hide'}
# Removing from the calculation the 3rd dataset of the classifier for the low number of samples included
huva.db_new <- huva.db
huva.db_new$classifier$data$PBMC.3 <- NULL
huva.db_new$classifier$anno$PBMC.3 <- NULL

# Selecting random genes
genes <- sample(rownames(huva.db_new$FG500$data$PBMC), 2000)

bench$gene <- vector()

for (i in 1:length(genes)) {
  tic()
  
  run_huva_experiment(data = huva.db_new, 
                      gene = genes[i], 
                      quantiles = 0.10, 
                      gs_list = hallmarks_V7.2,
                      summ = F, 
                      datasets_list = NULL, 
                      adjust.method = "none")
  
  time <- toc()
  time <- time$toc-time$tic
  bench$gene[i] <- time
}

names(bench$gene) <- genes
```

### *huva SOI experiment*
Here we will calculate the time required for the run of an *huva SOI experiment*
```{r, echo = T, results = 'hide'}
# Filtering the signatures for only those including > 100 genes
now_msig <- list()

for (i in names(msigdb_V7.2)) {
  
  if (length(msigdb_V7.2[[i]])>100) {
    
    now_msig[i] <- msigdb_V7.2[i]
    
  }
  
}

signature <- sample(now_msig, 500)

bench$signature <- vector()

for (i in 1:length(signature)) {
  tic()
  
  run_huva_signature(data = huva.db_new,   
                            gene_set = signature[[i]], 
                            GSVA.method =  "gsva",
                            quantiles = 0.10, 
                            gs_list = hallmarks_V7.2,
                            summ = F, 
                            datasets_list = NULL, 
                            adjust.method = "none")
  
  time <- toc()
  time <- time$toc-time$tic
  bench$signature[i] <- time
}

names(bench$signature) <- names(signature)
```

### *huva POI experiment*
Here we will calculate the time required for the run of an *huva POI experiment*
```{r, echo = T, results = 'hide'}
cell_count <- as.character(colnames(huva.db_new$FG500$metadata$cellcount))[-1]
cytokines <- as.character(colnames(huva.db_new$FG500$metadata$cytokines))[-1]

# Removing this parameter because of too many missing values
cytokines <- cytokines[cytokines!="IL6_Influenza_PBMC_24h"]

bench$phenotype <- vector()

for (i in 1:length(cell_count)) {
  tic()
  
  run_huva_phenotype(data = huva.db_new,
                     phenotype = cell_count[i],
                     study = "FG500",
                     metadata_table = "cellcount",
                     quantiles = 0.1, 
                     gs_list = hallmarks_V7.2)
  
  time <- toc()
  time <- time$toc-time$tic
  bench$phenotype[i] <- time
}

for (i in 1:length(cytokines)) {
  tic()
  
  run_huva_phenotype(data = huva.db_new,
                     phenotype = cytokines[i],
                     study = "FG500",
                     metadata_table = "cytokines",
                     quantiles = 0.1, 
                     gs_list = hallmarks_V7.2)
  
  time <- toc()
  time <- time$toc-time$tic
  bench$phenotype[length(cell_count)+i] <- time
}

names(bench$phenotype) <- c(cell_count, cytokines)
```

### *huva exploratory experiment*
Here we will calculate the time required for the run of an *huva exploratory experiment*
```{r, echo = T, results = 'hide'}
bench$explor <- vector()

for (i in genes) {
 
  tic()
  
  gene_exam(huva_dataset = huva.db_new, gene = i)
  
  time <- toc()
  time <- time$toc-time$tic
  bench$explor[i] <- time
}
```

### Plotting
#### Fig S2A
```{r}
plot(vector = bench$explor, title = "Fig. S2A - huva exploratory experiment", fill = "#4DBBD5", bins = 30)
dev.off()
```

#### Fig S2B
```{r}
plot(vector = bench$gene, title = "Fig. S2B - huva gene experiment", fill = "#E64B35", bins = 30)
dev.off()
```

#### Fig S2C
```{r}
plot(vector = bench$phenotype, title = "Fig. S2C - huva phenotype experiment", fill = "#01A087", bins = 30)
dev.off()
```

#### Fig S2D
```{r}
plot(vector = bench$signature, title = "Fig. S2D - huva signature experiment", fill = "#F39B7F", bins = 30)
dev.off()
```

## Extra - Plot the original results
You can from this chunk load the result shown in the manuscript:
```{r}
bench_orig <- readRDS("data/original_results.rds")
```

### Plotting

#### Fig S2A
```{r}
plot(vector = bench_orig$explor, title = "Fig. S2A - huva exploratory experiment", fill = "#4DBBD5", bins = 30)
dev.off()
```

#### Fig S2B
```{r}
plot(vector = bench_orig$gene, title = "Fig. S2B - huva gene experiment", fill = "#E64B35", bins = 30)
dev.off()
```

#### Fig S2C
```{r}
plot(vector = bench_orig$phenotype, title = "Fig. S2C - huva phenotype experiment", fill = "#01A087", bins = 30)
dev.off()
```

#### Fig S2D
```{r}
plot(vector = bench_orig$signature, title = "Fig. S2D - huva signature experiment", fill = "#F39B7F", bins = 30)
dev.off()
```

## Clean the env
```{r}
rm(progressBar, cell_count, cytokines, genes, i, iGeneSet, nGeneSets, time, signature)
```

# Session info
```{r}
info <- sessionInfo()
info
```

# Save environment
```{r}
save.image(paste("./data/",Sys.Date(), "_huva_Figure_2.RData", sep = ""))
```